Deciding Bitvector Termination with Program Synthesis

### 2014-07-05 papers/termination/paper.tex
formalisation

@@

All we need to do in order to use {\bf [CT]} and {\bf [CNT]} in any context %environment and regardless of the complexity of the loops's body 
is generate constraints abstracting 
the environment and the loop's body: for termination, we need over-approximating abstractions, whereas for non-termination 
we need under-approximations. 
As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
we will not provide the full algorithm, but rather illustrate it through examples.

- when analysing a loop, we assume that the nested ones and the one preceding it in program order terminate.\\

@@

\noindent {\bf Over-approximating the environment for termination.}
%% While proving that loop $L$ terminates, we create constraints on the environment in the
%% form of the predicate $I$ and the over-approximating summary relation $T_o$:
%% In the presence of nested loops, 
%% for each loop $L'$ nested in the body
%% of $L$, we must generate constraints saying that $T_o$ overapproximates $L'$, i.e. it is an over-approximation of the transitive closure of $L'$'s transition relation.
%% Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
%% must add constraints saying that $I$ over-approximates the states reachable via $L''$.
%The details of how to create these constraints are standard, so
%rather than presenting the algorithm for generating the constraints, we direct the
%reader's attention to Figure~\ref{fig:environment-model}.  
%
The example in Figure~\ref{fig:environment-model} contains three loops, which we will number 1 to 3 according to their program order.
Let us assume that we have already proven the termination of loops 1 and 3, and we are now trying to prove that loop 2 terminates. The 
termination of loop 2 is encoded as the conjunction of constraints on the right of Figure~\ref{fig:environment-model}:
the lines marked as $*$ over-approximate the states in which loop 2 can start execute.

contains a non-trivial program
along with the constraint system
encoding the termination of the outer loop.  Since the program contains sequential composition
of loops and nested loops, the constraint
system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.

Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
\begin{align*}
W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge z' + v' = x' \wedge v' \leq y' \\
R(x) & = x
\end{align*}

\begin{figure*}
\begin{framed}
\begin{minipage}{0.19\textwidth}
\begin{lstlisting}
x = 0;
while (x <= y)
  x++;

while (x > 0) {
  z = x;
  v = 0;

  while (v < y) {
    z--;
    v++;
  }

  x -= z;
  y -= z;
}
\end{lstlisting}
\end{minipage}
\vline
\begin{minipage}{0.82\textwidth}
\begin{align*}
 \exists W_1, W_2, R, S . \forall v, x, y, z, v', x', y', z' . \\
   x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
   x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
   x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x+1, y, z) \, \wedge \\
   %x > 0 \wedge W_2(v, x, y, z) & \rightarrow W_3(0, x, y, x) \, \wedge \\
   %v < y \wedge W_3(v, x, y, z) & \rightarrow W_3(v+1, x, y, z-1) \, \wedge \\
   %v \geq y \wedge W_3(v, x, y, z) & \rightarrow W_2(v, x-z, y-z, z) \, \wedge \\
   S(\langle v, x, y, z, \rangle, \langle v', x', y', z' \rangle) \wedge v < y & \rightarrow S(\langle v, x, y, z \rangle, \langle v+1, x, y, z-1 \rangle) \, \wedge \\
   x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
   x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
   & W_2(v', x' - z', y' - z', z') \, \wedge \\
   & R(v, x, y, z) > 0 \, \wedge \\
   & R(v, x, y, z) > R(v', x' - z', y'-z', z')
\end{align*}
\end{minipage}
\caption{A non-trivial program and its constraint system\label{fig:environment-model}}
\end{framed}
\end{figure*}

\noindent {\bf Under-approximating the environment for non-termination.}
Dual to termination, when attempting to prove non-termination, we must under-approximate the environment.

@@

%% \subsection{Program Synthesis as Second-Order Satisfaction}

>>>>

%\item We propose a non-trivial formulation of the (non-)termination problem inside a second order logic fragment with only one quantifier alternation.

@@

All we need to do in order to use {\bf [CT]} and {\bf [CNT]} in any context %environment and regardless of the complexity of the loops's body 
is generate constraints abstracting 
the environment and the loop's body: for termination, we need over-approximating abstractions, whereas for non-termination 
we need under-approximations. 
As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
we will not provide the full algorithm, but rather illustrate it through examples.\\

@@

\noindent {\bf Over-approximating the environment for termination.}
%% While proving that loop $L$ terminates, we create constraints on the environment in the
%% form of the predicate $I$ and the over-approximating summary relation $T_o$:
%% In the presence of nested loops, 
%% for each loop $L'$ nested in the body
%% of $L$, we must generate constraints saying that $T_o$ overapproximates $L'$, i.e. it is an over-approximation of the transitive closure of $L'$'s transition relation.
%% Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
%% must add constraints saying that $I$ over-approximates the states reachable via $L''$.
%The details of how to create these constraints are standard, so
%rather than presenting the algorithm for generating the constraints, we direct the
%reader's attention to Figure~\ref{fig:environment-model}.  
%
The example in Figure~\ref{fig:environment-model} contains three loops. %, which we will number 1 to 3 according to their program order.
Let us assume that we have already proven the termination of $L_1$ and $L_3$, and we are now trying to prove that $L_2$ terminates, which we encode
as the conjunction of constraints on the right side of the figure:
\begin{itemize}
\item $W_1$, $W_2$ and $W_3$ are (over-approximating) inductive invariants for $L_1$, $L_2$ and $L_3$, respectively.
\item $S$ is a summary of $L_3$ that over-approximates the transitive closure of its transition relation.
\item If the formula is satisfiable, then $R$ is a ranking function for $L_2$.
 \end{itemize}

%the lines marked as $*$ over-approximate the states in which loop 2 can start execute.
%% contains a non-trivial program
%% along with the constraint system
%% encoding the termination of the outer loop.  Since the program contains sequential composition
%% of loops and nested loops, the constraint
%% system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.

Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
\begin{align*}
W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge \\
& z' + v' = x' \wedge v' \leq y' \\
R(x) & = x
\end{align*}

\begin{figure*}
\begin{framed}
\begin{minipage}{0.22\textwidth}
\begin{lstlisting}[mathescape=true]
x = 0;
$L_1:$ while (x <= y)
  x++;

$L_2:$ while (x > 0) {
  z = x;
  v = 0;

$L_3:$  while (v < y) {
    z--;
    v++;
  }

  x -= z;
  y -= z;
}
\end{lstlisting}
\end{minipage}
\vline
\begin{minipage}{0.82\textwidth}
\begin{align*}
 \exists W_1, W_2, R, S . \forall v, x, y, z, v', x', y', z' . \\
   x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
   x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
   x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x+1, y, z) \, \wedge \\
   %x > 0 \wedge W_2(v, x, y, z) & \rightarrow W_3(0, x, y, x) \, \wedge \\
   %v < y \wedge W_3(v, x, y, z) & \rightarrow W_3(v+1, x, y, z-1) \, \wedge \\
   %v \geq y \wedge W_3(v, x, y, z) & \rightarrow W_2(v, x-z, y-z, z) \, \wedge \\
   S(\langle v, x, y, z, \rangle, \langle v', x', y', z' \rangle) \wedge v < y & \rightarrow S(\langle v, x, y, z \rangle, \langle v+1, x, y, z-1 \rangle) \, \wedge \\
   x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
   x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
   & W_2(v', x' - z', y' - z', z') \, \wedge \\
   & R(v, x, y, z) > 0 \, \wedge \\
   & R(v, x, y, z) > R(v', x' - z', y'-z', z')
\end{align*}
\end{minipage}
\end{framed}
\caption{A non-trivial program and its constraint system\label{fig:environment-model}}
\end{figure*}

\noindent {\bf Under-approximating the environment for non-termination.}
Dual to termination, when attempting to prove non-termination, we must under-approximate the environment and the loop body's and apply 
formula {\bf [CNT]}. However, there is one more constraint that we must obey when doing so: we must stay inside the synthesis fragment! 
This can be achieved for the environment as its under-approximation $I$ and the loop under consideration are at the same level of nestedness and do not therefore 
require another quantifier alternation.

However, when dealing with nested loops, computing the under-approximation of the loop's body 
requires an extra quantifier alternation. The origin of the problem comes from the fact that an under-approximation is an existential property and therefore
requires an extra existential quantifier, resulting in the $\exists \forall \exists$ alternation. This situation is similar to the one encountered when proving non-termination of non-deterministic loops. 
To go around this limitation, we treat nested loops as a special case and give a separate formula for them in Figure~\ref{fig:nonterm-nested}. If the formula on the right of the figure is satisfiable, then $L_1$ is non-terminating as witnessed by the recurrence set $N_1$ and
the initial state $x_0$ in which the program begins executing. There are two possible scenarios for $L_2$'s termination:
\begin{itemize}
\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x) \wedge P_2(x,x') \rightarrow N_1(x') $. 
\item If $L_2$ is non-terminating, then $N_2$ is its recurrence set.
\end{itemize}

%% The issue appears when we try to under-approximate the body of the loop. 
%% Here we encounter a similar problem with the one for non-determinism mentioned above.

%% Consider the generic example in Figure~\ref{fig:nonterm-environment}, where we are attempting to prove non-termination
%% of the innermost loop $L_3$ by finding a closed recurrence set $N_3$. 
%% -- We will assume that the other two loops iterate only a finite number of times.

%% We now explain the constraints on the right of the figure.
%% Given that we prove non-termination, we try to  under-approximate the states from which $L_3$ can start executing. 
%% We do this in several steps:

%% $W_1$, $W_2$ and $W_3$ are inductive invariants corresponding to $L_1$, $L_2$ and $L_3$, respectively, such that:

%% - $W_1$ is an inductive invariant of $L_1$ that is strong enough to establish $W_2$ if $L_1$ exits (which it does by assumption).
%% - $W_2$ is an inductive invariant of $L_2$ that is strong enough to establish either $W_3$ or $N_3$ on entry to $L_3$, depending on whether $L_3$ terminates or not.
%% - Recall that $N_3$ is a recurrence set for $L_3$, so if a state $x \in N_3$ is reached, the loop $L_3$ will execute forever. Otherwise, $W_3$ holds on entry to $L_3$.
%% - $W_3$ is an inductive invariant of $L_3$ that is strong enough to re-establish $W_2$ when $L_3$ exits.

%% If the non-termination formula is satisfiable, $x_0$ is some initial state in which the program begins executing.

%% First, we under-approximate the states in which $L_2$ can start executing through $W_2$.
%% Then $N_1$ is an inductive invariant of $L_1$ that is strong
%% enough to establish $N_2$ if $L_1$ exits (which it does by assumption).  Now, $N_2$ is an inductive invariant
%% of $L_2$ that is strong enough to establish either $N_3$ or $N_4$ on entry to $L_3$.  Recall that $N_4$ is
%% a recurrence set for $L_3$, so if a state $x \in N_4$ is reached, the loop $L_3$ will execute forever.
%% Otherwise, $N_3$ holds on entry to $L_3$.  $N_3$ is an inductive invariant of $L_3$ that is strong enough to
%% re-establish $N_2$ when $L_3$ exits.

%% If the constraint system is satisfiable, $x_0$ is some initial
%% state in which the program begins executing.

%%   Since we assumed that $L_2$ iterated only a finite number of times,
%% $N_3$ must be established only a finite number of times, which means that $N_4$ will eventually be reached,
%% at which point $L_3$ will loop forever since $N_4$ is a recurrence set for $L_3$.

%% \begin{figure*}
%% \begin{framed}
%%  \begin{minipage}{0.19\textwidth}
%% \begin{lstlisting}
%% L1: while (c1) {
%%      B1;
%%     }

%% L2: while (c2) {
%%      P;

%% L3:  while (c3) {
%%        B3;
%%      }
%%     }
%% \end{lstlisting}
%% \end{minipage}
%% \vline
%% \begin{minipage}{0.82\textwidth}
%% \begin{align*}
%%  \exists W_1, W_2, W_3, N_3, x_0 . \forall x, x' . \\
%%   W_1(x_0) \, \wedge \\
%%   c_1(x) \wedge W_1(x) \wedge B_1(x, x') & \rightarrow W_1(x') \, \wedge \\
%%   \lnot c_1(x) \wedge W_1(x) & \rightarrow W_2(x) \, \wedge \\
%%   W_2(x) & \rightarrow c_2(x) \, \wedge \\
%%   W_2(x) \wedge P(x, x') & \rightarrow W_3(x') \vee N_3(x') \, \wedge \\
%%   c_3(x) \wedge W_3(x) \wedge B_3(x, x') & \rightarrow W_3(x') \, \wedge \\
%%   \lnot c_3(x) \wedge W_3(x) & \rightarrow W_2(x) \, \wedge \\
%%   N_3(x) & \rightarrow c_3(x) \, \wedge \\
%%   N_3(x) \wedge B_3(x, x') & \rightarrow N_3(x')
%% \end{align*}
%% \end{minipage}
%% \end{framed}

%% \caption{The constraint system for non-termination\label{fig:nonterm-environment}}
%% \end{figure*}

\begin{figure*}
\begin{framed}
 \begin{minipage}{0.19\textwidth}
\begin{lstlisting}[mathescape=true]
$L_1$: while ($G_1$) {
     $P_1$;

$L_2$:  while ($G_2$) {
       $B_2$;
       }

     $P_2$;
     }
\end{lstlisting}
\end{minipage}
\vline
\begin{minipage}{0.82\textwidth}
\begin{align*}
 \exists N_1, N_2, x_0 . \forall x, x' . \\
  N_1(x_0) & \, \wedge \\
  N_1(x) & \rightarrow G_1(x) \, \wedge \\
  N_1(x) \wedge P_1(x,x') & \rightarrow N_2(x') \, \wedge \\
  G_2(x) \wedge N_2(x) \wedge B_2(x,x') & \rightarrow N_2(x') \, \wedge \\
  \lnot G_2(x) \wedge N_2(x) \wedge P_2(x,x') & \rightarrow N_1(x') 
\end{align*}
\end{minipage}
\end{framed}

\caption{The non-termination formula for nested loops \label{fig:nonterm-nested}}
\end{figure*}

@@

- when analysing a loop, we assume that the nested ones and the one preceding it in program order terminate.\\
%% \subsection{Program Synthesis as Second-Order Satisfaction}
